Surface correlation functions calculations: improving methodology for discrete images with novel exact continuous approach

### 2023-03-23 paper.tex
Add an algorithm for F_{sss}

\section{Algorithm for computation of surface functions}
\label{sec:algo}
Here we briefly describe an algorithm for computation of surface-surface and
surface-void correlation functions for digital images of porous media. A more
detailed description can be found in a paper of Samarin
et~al. \textcolor{red}{cite paper} This algorithm can calculate surface
correlation functions for all possible directions and correlation lengths. It
consists of two stages: edge detection stage and computation of
autocorrelation.

We give an image of porous medium in the form of two- or three-dimensional bit
array $A$ as the input to our algorithm. In the first stage, we convolve the
input with high-pass filter to extract edges from the image. We use the
following filter kernel $H$ \textcolor{red}{Is there notation for 3x3x3
  ``matrices''?}:
\begin{align}
  H_{ij} &= \frac{1}{6} \left[
    \begin{array}{ccc}
      1 & 1 & 1 \\
      1 & -8 & 1 \\
      1 & 1 & 1
    \end{array}
    \right] \label{eq:filter-3x3-2d} \\
  H_{ijk} &= \frac{1}{18} \left[
    \begin{array}{ccc}
      1 & 1 & 1 \\
      1 & 1 & 1 \\
      1 & 1 & 1
    \end{array} ;
    \begin{array}{ccc}
      1 & 1 & 1 \\
      1 & -24 & 1 \\
      1 & 1 & 1
    \end{array} ;
    \begin{array}{ccc}
      1 & 1 & 1 \\
      1 & 1 & 1 \\
      1 & 1 & 1
    \end{array}
    \right] \label{eq:filter-3x3-3d}
\end{align}
Here symbol $;$ denotes concatenation along the third dimension. Edge extraction
is performed as follows:
\begin{align*}
  A'_{lm}  &= \left| \sum_i\sum_j A_{l-i, m-j}H_{ij} \right| \quad \text{for 2D} \\
  A'_{lmn} &= \left| \sum_i\sum_j\sum_k A_{l-i, m-j, n-k}H_{ijk} \right| \quad
  \text{for 3D}
\end{align*}
Therefore $A'$ has non-zero elements only in a small area around interface
between phases in $A$.

In the second stage we calculate autocorrelation for $A'$ and divide it by a
total number of elements in $A$:
\begin{align*}
  F_{{ss}_{lm}} &= \frac{1}{N} \sum_i \sum_j A'_{ij}A'_{i+l, j+m} \quad
  \text{for 2D} \\
  F_{{ss}_{lmn}} &= \frac{1}{N} \sum_i \sum_j \sum_k A'_{ijk}A'_{i+l, j+m, k+n}
  \quad \text{for 3D}
\end{align*}
Calculation of sums in the latter equations can be performed using the following
properties of Fourier transform:
\begin{equation*}
  F[f(x) \star f(x)] = F[f(x) * f(-x)] = \hat{f}(z)\overline{\hat{f}(z)} = |\hat{f}(z)|^2
\end{equation*}

@@

From now on we will work with square images with dimensions $1 \times 1$ square
length units, so we can omit division by $N$ in described algorithms and
ensemble average in \cref{eq:fss} and \cref{eq:fsv} becomes a double integral
over a square $[0, 1]^2$.

>>>>

\section{Algorithm for computation of surface functions}
\label{sec:algo}
Here we briefly describe an algorithm for computation of surface-surface,
surface-void and surface-surface-surface correlation functions for digital
images of porous media. A more detailed description can be found in a paper of
Samarin et~al. \textcolor{red}{cite paper} This algorithm can calculate surface
correlation functions for all possible directions and correlation lengths. It
consists of two stages: edge detection stage and computation of autocorrelation.

We give an image of porous medium in the form of two- or three-dimensional bit
array $A$ as the input to our algorithm. In the first stage, we convolve the
input with high-pass filter to extract edges from the image. We use the
following filter kernel $H$ \textcolor{red}{Is there notation for 3x3x3
  ``matrices''?}:
\begin{align}
  H_{ij} &= \frac{1}{6} \left[
    \begin{array}{ccc}
      1 & 1 & 1 \\
      1 & -8 & 1 \\
      1 & 1 & 1
    \end{array}
    \right] \label{eq:filter-3x3-2d} \\
  H_{ijk} &= \frac{1}{18} \left[
    \begin{array}{ccc}
      1 & 1 & 1 \\
      1 & 1 & 1 \\
      1 & 1 & 1
    \end{array} ;
    \begin{array}{ccc}
      1 & 1 & 1 \\
      1 & -24 & 1 \\
      1 & 1 & 1
    \end{array} ;
    \begin{array}{ccc}
      1 & 1 & 1 \\
      1 & 1 & 1 \\
      1 & 1 & 1
    \end{array}
    \right] \label{eq:filter-3x3-3d}
\end{align}
Here symbol $;$ denotes concatenation along the third dimension. Edge extraction
is performed as follows:
\begin{align*}
  A'_{lm}  &= \left| \sum_i\sum_j A_{l-i, m-j}H_{ij} \right| \quad \text{for 2D} \\
  A'_{lmn} &= \left| \sum_i\sum_j\sum_k A_{l-i, m-j, n-k}H_{ijk} \right| \quad
  \text{for 3D}
\end{align*}
Therefore $A'$ has non-zero elements only in a small area around interface
between phases in $A$. This operation is convolution of a filter $H$ with an
image $A$ and can be performed using Fourier transform:
\begin{equation*}
  A' = H*A = F^{-1}[F[H] F[A]]
\end{equation*}
Here $F$ is a forward Fourier transform and $F^{-1}$ is its inverse.

In the second stage we calculate autocorrelation for $A'$ and divide it by a
total number of elements in $A$:
\begin{align*}
  F_{{ss}_{lm}} &= \frac{1}{N} \sum_i \sum_j A'_{ij}A'_{i+l, j+m} \quad
  \text{for 2D} \\
  F_{{ss}_{lmn}} &= \frac{1}{N} \sum_i \sum_j \sum_k A'_{ijk}A'_{i+l, j+m, k+n}
  \quad \text{for 3D}
\end{align*}
Calculation of sums in the latter equations can be performed using the following
properties of Fourier transform (here $\hat{f}(z) = F[f(x)]$):
\begin{equation*}
  F[f(x) \star f(x)] = F[f(x) * f(-x)] = \hat{f}(z)\overline{\hat{f}(z)} = |\hat{f}(z)|^2
\end{equation*}

@@

For surface-surface-surface function $F_{sss}$ the former approach can also be
used, but it would be highly inefficient memory-wise because for three
dimensional case $F_{sss}$ is a function of six scalar coordinates which leads
to a very large output array. Instead of this approach we use a slower approach
which computes $F_{sss}$ at a specified point calculating a convolution using
its definition:
\begin{algorithmic}[1]
  \Procedure{surfsurfsurf}{$A$, $\bm{r_1}$, $\bm{r_2}$}
  \State $A' \gets A*H$
  \Comment{$A'$ becomes an interface of $A$.}
  \State $A'_1 \gets shift(A', \bm{r_1})$
  \Comment{Shift $A$ in $i$-th dimension by $r_{1i}$ for all $i \in \overline{1, ndim(A)}$}
  \State $A'_2 \gets shift(A', \bm{r_2})$
  \Comment{Shift $A$ in $i$-th dimension by $r_{2i}$ for all $i \in \overline{1, ndim(A)}$}
  \State $M \gets A' \cdot A'_1 \cdot A'_2$
  \Comment{Multiplicate three arrays element-wise}
  \State \textbf{return} $\sum_{i} M_{i} / N$
  \Comment{Sum all elements in $M$ and divide by a number of elements in $A$}
  \EndProcedure
\end{algorithmic}
This algorithm can be vectorized and therefore benefits from the usage of
GPUs. Shift function in this algorithm must be a circular shift which
``rotates'' content of the array, if periodic boundary conditions are
desired. For our purpose we use $F_{sss}$ function in the form
$F_{sss}(r_1, r_2, \theta)$ and fix $\theta = \pi/2$. This leads us to a
sampling pattern shown on \cref{fig:Fsss-pattern}
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/pattern.png}
  \caption[]{Sampling pattern used for calculation of $F_{sss}$ in our
    implementation is a right triangle with varying catets $r_1$ and
    $r_2$. Patterns can lie in planes $XY$ (as on the picture), $XZ$ or
    $YZ$. Black dots depict elements of $A'$}
  \label{fig:Fsss-pattern}
\end{figure}

From now on when considering $F_{ss}$ and $F_{sv}$ functions we will work with
square images which have unit area, so we can omit division by $N$ in described
algorithms and ensemble average in \cref{eq:fss} and \cref{eq:fsv} becomes a
double integral over a square $[0, 1]^2$. When considering $F_{sss}$ function we
will work with cubic images which have unit volume. Then we can also omit
division by $N$ and ensemble average in \cref{eq:fsss} becomes a triple integral
over a cube $[0, 1]^3$.
